---
title: "Class 11"
author: "Christeena Pano (PID: A17842498)"
date: 05/07/2026
format: pdf
---

## Custom analysis of resulting models

Read the results of the more complicated HIV-Pr dimer AlphaFold2 models into R with the help of the Bio3D package

```{r}
# Change this for YOUR results dir name
results_dir <- "hivprdimer_23119/" 
```

```{r}
# File names for all PDB models
pdb_files <- list.files(path=results_dir,
                        pattern="*.pdb",
                        full.names = TRUE)

# Print our PDB file names
basename(pdb_files)
```

```{r}
library(bio3d)

# Read all data from Models 
#  and superpose/fit coords
pdbs <- pdbaln(pdb_files, fit=TRUE, exefile="msa")
```

```{r}
pdbs
```

```{r}
# RMSD is a standard measure of structural distance between coordinate sets. Use the rmsd() function to calculate the RMSD between all pairs models.
rd <- rmsd(pdbs, fit=T)

range(rd)
```

```{r}
# Draw a heatmap of these RMSD matrix values

library(pheatmap)

colnames(rd) <- paste0("m",1:5)
rownames(rd) <- paste0("m",1:5)
pheatmap(rd)
```

```{r}
# Read a reference PDB structure
pdb <- read.pdb("1hsg")
```

```{r}
# Obtain secondary structure from a call to stride() or dssp() on any of the model structures
plotb3(pdbs$b[1,], typ="l", lwd=2, sse=pdb)
points(pdbs$b[2,], typ="l", col="red")
points(pdbs$b[3,], typ="l", col="blue")
points(pdbs$b[4,], typ="l", col="darkgreen")
points(pdbs$b[5,], typ="l", col="orange")
abline(v=100, col="gray")
```

```{r}
# Use the core.find() function to find the most consistent “rigid core” common across all the models
core <- core.find(pdbs)
```

```{r}
# Use the identified core atom positions as a basis for a superposition and write out fitted structures to a directory called corefit_structure
core.inds <- print(core, vol=0.5)
```

```{r}
xyz <- pdbfit(pdbs, core.inds, outpath="corefit_structures")
```

RMSF is an often used measure of conformational variance along the structure

```{r}
# Explore RMSF between positions of the structure
rf <- rmsf(xyz)

plotb3(rf, sse=pdb)
abline(v=100, col="gray", ylab="RMSF")
```

Independent of the 3D structure, AlphaFold produces an output called Predicted Aligned Error (PAE). This is detailed in the JSON format result files, one for each model structure.

```{r}
# read these files and see that AlphaFold produces a useful inter-domain prediction for models 1 and 2.
library(jsonlite)

# Listing of all PAE JSON files
pae_files <- list.files(path=results_dir,
                        pattern=".*model.*\\.json",
                        full.names = TRUE)
```

```{r}
# Read the 1st-5th file
pae1 <- read_json(pae_files[1],simplifyVector = TRUE)
pae2 <- read_json(pae_files[2],simplifyVector = TRUE)
pae3 <- read_json(pae_files[3],simplifyVector = TRUE)
pae4 <- read_json(pae_files[4],simplifyVector = TRUE)
pae5 <- read_json(pae_files[5],simplifyVector = TRUE)

attributes(pae1)
```

```{r}
# Per-residue pLDDT scores 
#  same as B-factor of PDB..
head(pae1$plddt) 
```

```{r}
# Find max PAE scores
pae1$max_pae
pae2$max_pae
pae3$max_pae
pae4$max_pae
pae5$max_pae
```
We can plot the N by N (where N is the number of residues) PAE scores with ggplot or with functions from the Bio3D package

```{r}
# Plot the N by N PAE scores
plot.dmat(pae1$pae, 
          xlab="Residue Position (i)",
          ylab="Residue Position (j)")
```

```{r}
plot.dmat(pae1$pae, 
          xlab="Residue Position (i)",
          ylab="Residue Position (j)",
          grid.col = "black",
          zlim=c(0,30))
```

```{r}
plot.dmat(pae2$pae, 
          xlab="Residue Position (i)",
          ylab="Residue Position (j)",
          grid.col = "black",
          zlim=c(0,30))
```

```{r}
plot.dmat(pae3$pae, 
          xlab="Residue Position (i)",
          ylab="Residue Position (j)",
          grid.col = "black",
          zlim=c(0,30))
```

```{r}
plot.dmat(pae4$pae, 
          xlab="Residue Position (i)",
          ylab="Residue Position (j)",
          grid.col = "black",
          zlim=c(0,30))
```

```{r}
plot.dmat(pae5$pae, 
          xlab="Residue Position (i)",
          ylab="Residue Position (j)",
          grid.col = "black",
          zlim=c(0,30))
```

## Residue conservation from alignment file

```{r}
# Read alignment file from Alphafold2 results to analyze residue conservation
aln_file <- list.files(path=results_dir,
                       pattern=".a3m$",
                        full.names = TRUE)
aln_file
```

```{r}
# Read first alignment file as fasta object
aln <- read.fasta(aln_file[1], to.upper = TRUE)
```

```{r}
# Check number of sequences in the alignment
dim(aln$ali)
```

```{r}
# Use the conserv() function to score residue conservation in the alignment
sim <- conserv(aln)

# Plot results 
plotb3(sim[1:99], sse=trim.pdb(pdb, chain="A"),
       ylab="Conservation Score")
```

```{r}
# Generate consensus sequence with high cutoff value
con <- consensus(aln, cutoff = 0.9)
con$seq
```

we can map this conservation score to the Occupancy column of a PDB file for viewing in molecular viewer programs such as Mol*

```{r}
m1.pdb <- read.pdb(pdb_files[1])
occ <- vec2resno(c(sim[1:99], sim[1:99]), m1.pdb$atom$resno)
write.pdb(m1.pdb, o=occ, file="m1_conserv.pdb")
```



