---
title: "Homework 6"
author: "Christeena Pano (PID: A7842497)"
format: pdf
---

```{r}
library(bio3d) 
```

## Comments from rubric

> *Inputs to the function*: The protein ID from the Protein Data Bank (PDB)

> *What the function does and how to use it*: This function uses the protein ID to analyze proteins by trimming them to a specific chain (A) and atom/element type (CA), and reading their b-factor values. This is used to generate a plot from any PDB ID, with the specific characteristics wanted. This function is used by using the function `pdb.plot` and using the `PDB ID` as the input. 

> *Output of the function*: The output of the function is a b-factor plot as well as the b-factor data in the form of a vector. 

```{r}
pdb.plot <- function(pdb_id) {

  structure <- read.pdb(pdb_id)
  
  structure.trim <- trim.pdb(structure, chain = "A", elety = "CA")
  
  b <- structure.trim$atom$b
  
  plotb3(b, sse = structure.trim, typ = "l", ylab = "B-factor")
  
  return(b)
}
```

## Function works and correct output is produced

```{r}
pdb.plot("4AKE")
```



